---
title: "P8105 HW2"
author: "Lucia Wang (lw3061)"
date: "due 10-04-2023"
output: github_document
---
```{r}
library(tidyverse)
library(readxl)
```

# Problem 1
## Merge some FiveThirtyEight datasets together.

1) clean the data in `pols-month.csv`
```{r}
pols_month_df = 
  read_csv("fivethirtyeight_datasets/pols-month.csv") |>
  janitor::clean_names() |>
  separate(mon, c("year", "month", "day")) |>
  mutate(
    month = case_match(month,
                       "10" ~ "October",
                      "11" ~ "November",
                       "12" ~ "December",
                      "01" ~ "January",
                      "02" ~ "February",
                      "03" ~ "March",
                      "04" ~ "April",
                      "05" ~ "May",
                      "06" ~ "June",
                      "07" ~ "July",
                      "08" ~ "August",
                      "9" ~ "September"),
    president = case_match(prez_dem, 
                           1 ~ "dem",
                           0 ~ "gop"),
    year = as.numeric(year)
  ) |>
  select(c(-prez_dem, -prez_gop, -day)) |> 
  arrange(year, month)
```

2) clean the data in `snp.csv` (solution adapted from posted solutions)
```{r}
snp_df = 
  read_csv("fivethirtyeight_datasets/snp.csv") |>
  separate(date, into = c("month", "day", "year"), convert = TRUE) |>
  arrange(year, month) |>
  mutate(month = month.name[month]) |>
  select(year, month, close) 
```

3) do the same with `unemployment` (solution adapted from posted solutions)
```{r}
months = 
  tibble(
    month_num = 1:12,
    month_abb = month.abb,
    month = month.name
  )

unemployment_df = 
  read_csv("fivethirtyeight_datasets/unemployment.csv") |>
  rename(year = Year) |>
  pivot_longer(
    Jan:Dec, 
    names_to = "month_abb",
    values_to = "unemployment"
  ) |> 
  left_join(x = _, y = months) |> 
  select(year, month, unemployment)
```

4) join datasets by merging (solution adapted from posted solutions)
```{r}
merge538 = 
  left_join(pols_month_df, snp_df) |>
  left_join(x = _, y = unemployment_df)
```

The dataset `pols_month_df` contained data about the political parties in office at specific times, with `r nrow(pols_month_df)` observations and `r ncol(pols_month_df)` variables. `snp_df` had `r nrow(snp_df)`observations and `r ncol(snp_df)` variables, and `unemployment_df` had `r nrow(unemployment_df)` observations and `r ncol(unemployment_df)` variables. 

The final merged dataset gives information about the president, governors, representatives, and senators in office at specific times between 1945 and 2015, as well as information about unemployment rates. 

# Problem 2
### Read and clean the Mr. Trash Wheel sheet.

1) specify the sheet and omit rows with notes/figures and columns with notes
```{r}
mrtrash_df = 
  read_excel("202309 Trash Wheel Collection Data.xlsx", 
             sheet="Mr. Trash Wheel", 
             range="A2:M586")
```

2) clean and update the data to include a new `homes_powered` variable. Add an additional variable to keep track of which Trash Wheel is which...
```{r}
mrtrash_df =
  janitor::clean_names(mrtrash_df) |>
  mutate(kilowatts = weight_tons * 500,
         homes_powered = kilowatts / 30,
         name = "Mr. Trash Wheel") 

mrtrash_df
```

3) repeat for Professor Trash Wheel and Gwynnda, then combine
```{r}
proftrash_df =
  read_excel("202309 Trash Wheel Collection Data.xlsx", 
             sheet="Professor Trash Wheel", 
             range="A2:L108") |>
  janitor::clean_names() |>
  mutate(kilowatts = weight_tons * 500,
         homes_powered = kilowatts / 30,
         name = "Professor Trash Wheel",
         year = as.character(year)) 

proftrash_df

gwynnda_df = 
  read_excel("202309 Trash Wheel Collection Data.xlsx", 
             sheet="Gwynnda Trash Wheel", 
             range="A2:K157") |>
  janitor::clean_names() |>
  mutate(kilowatts = weight_tons * 500,
         homes_powered = kilowatts / 30,
         name = "Gwynnda Trash Wheel",
         year = as.character(year))

gwynnda_df

trashwheels_tidy_df =
  bind_rows(mrtrash_df, proftrash_df, gwynnda_df) |>
  janitor::clean_names()

```

The final tidied dataset with all 3 trash wheels incorporated has `r ncol(trashwheels_tidy_df)` columns (variables) and `r nrow(trashwheels_tidy_df)` rows (observations). 

Key variables are `weight_tons`, `homes_powered`, `dumpster`, and `month`/`year`/`date`. Some of the most popular trash items were `cigarette_butts`, `plastic_bottles`, and `polystyrene`, but others like `sports_balls` and `plastic_bags` were also present. 

The total weight of trash collected by Professor Trash Wheel was `r sum(pull(proftrash_df, weight_tons))` tons. 

The total number of cigarette butts collected by Gwynnda in July 2021 was `r sum(pull(filter(gwynnda_df, month=="July", year=="2021"),cigarette_butts))`.


# Problem 3
### Import and clean MCI data.

1) import the data and clean names
```{r}
baseline_df = 
  read_csv("data_mci/MCI_baseline.csv", skip=1) |>
  janitor::clean_names() 
```
The data currently has `r nrow(baseline_df)` observations which is the number of participants that were recruited.

2) recode variables and remove participants who did not have an age of onset of MCI
```{r}
baseline_df = baseline_df |>
  mutate(
    sex = as.character(sex),
    sex = case_match(sex, 
                     "0" ~ "Female",
                     "1" ~ "Male"),
    apoe4 = as.character(apoe4),
    apoe4 = case_match(apoe4,
                       "0" ~ "Non-carrier",
                       "1" ~ "Carrier"),
    id = as.character(id)
        ) |>
  filter(age_at_onset == "." | age_at_onset >= current_age) 
```
The number of participants that developed MCI was `r nrow(filter(baseline_df, age_at_onset != "."))`. 

The average baseline age of this final dataset was `r round(mean(pull(baseline_df,current_age)),digits=2)` years.

The proportion of women in the study who were APOE4 carriers was `r round(nrow(filter(baseline_df, sex=="Female", apoe4=="Carrier"))/nrow(filter(baseline_df, sex=="Female")),digits=2)`.


3) import, clean, and tidy the biomarkers csv
```{r}
amyloid_df = 
  read_csv("data_mci/mci_amyloid.csv", skip=1) |>
  janitor::clean_names()
```
Before cleaning out missing time values, the total number of observations in the amyloid dataset is `r nrow(amyloid_df)`. 

```{r}
amyloid_df = amyloid_df |>
  drop_na() |>
  rename(id=study_id) |>
  mutate(
    id = as.character(id)
  )
```
After dropping missing values, the number of observations is `r nrow(amyloid_df)`.

4) see which participants appear in which datasets, and then combine them and only retain participants in both. 
```{r}
anti_combine_1 = 
  anti_join(baseline_df, amyloid_df, by="id")

anti_combine_2 = 
  anti_join(amyloid_df, baseline_df, by="id")
```
`r nrow(anti_combine_1)` participants only appeared in the baseline dataset, and `r nrow(anti_combine_2)` participants only appeared in the amyloid dataset. 

```{r}
mci_amyloid_combine = 
  inner_join(baseline_df, amyloid_df, by="id")
```
The final tidied dataset combining baseline and amyloid data only had full data for `r nrow(mci_amyloid_combine)` participants.The average age at baseline was `r round(mean(pull(mci_amyloid_combine, current_age)), digits=2)` years.

```{r, warning=FALSE, echo=FALSE, message=FALSE}
developed_mci = mci_amyloid_combine |>
  filter(age_at_onset != ".") |>
  mutate(
    age_at_onset = as.numeric(age_at_onset)
  )
```

For the `r nrow(filter(mci_amyloid_combine, age_at_onset != "."))` participants who developed MCI who had biomarker values, the average age of onset was `r round(mean(pull(developed_mci, age_at_onset)), digits=2)` years.

Of all participants, `r nrow(filter(mci_amyloid_combine, sex=="Female"))` were women. `r nrow(filter(developed_mci, sex=="Female"))` of subjects that developed MCI were women. 

The proportion of participants who were carriers of the APOE4 variant was `r round(nrow(filter(mci_amyloid_combine, apoe4=="Carrier")) / nrow(mci_amyloid_combine), digits=2)`. The proportion in those who developed MCI was `r round(nrow(filter(developed_mci, apoe4=="Carrier")) / nrow(developed_mci), digits=2)`. 

5) Export the data to the directory.
```{r, message=FALSE}
write_csv(mci_amyloid_combine, "data_mci/combined_mci_amyloid.csv")
```
